---
title: "Methodology: framework, definitions and analyses"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Methodology: framework, definitions and analyses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", out.width = "100%")
library(chokepointR)
```

This vignette documents the conceptual framework, the definitions of every
measure, and the analyses behind `chokepointR`, so that results can be cited and
reproduced in academic work. It is written to be read alongside the data
dictionaries (`?chokepoint_resilience`, `?chokepoint_context`,
`?chokepoint_sources`) and the provenance notes in `DATA-PROVENANCE.md`.

## 1. Motivation and scope

A large share of world seaborne trade -- and most seaborne oil -- must pass
through a small number of maritime **chokepoints**: narrow straits and
artificial canals with no close substitute. Because flows concentrate there, a
localised disruption (grounding, drought, conflict, closure) can propagate
through global value chains far beyond the immediate region. This concentration
is precisely why chokepoints are treated as systemic nodes in the economic-
geography and trade-resilience literatures [@verschuur2025; @unctad2024].

`chokepointR` covers **eight** maritime chokepoints chosen for their combination
of traffic volume and strategic irreplaceability: the Panama and Suez Canals,
and the Straits of Hormuz, Malacca, Bab el-Mandeb, Gibraltar, Dover and the
Turkish Straits. The package is deliberately **maritime and chokepoint-scoped**:
port clusters and inland networks (e.g. Black Sea ports, the Mississippi system)
appear in source incidents but are out of scope for the profile, so that every
profiled unit is comparable.

The design principle throughout is **transparency over authority**: the package
ships facts with per-cell attribution and a reproducible index recipe, not a
black-box score. Every component is exposed so users can inspect, re-weight or
discard it.

## 2. The risk taxonomy

Incidents are coded with a fixed **11-risk, 3-category** taxonomy
(`risk_types`), independent of which risks happen to be observed in any given
period:

```{r taxonomy, echo = FALSE, results = "asis"}
knitr::kable(cp_risk()[order(cp_risk()$risk_category), ],
             row.names = FALSE, col.names = c("Risk", "Category", "Code"))
```

The three categories -- **weather and climate**, **security and conflict**, and
**political and institutional** -- separate the *mechanism* of disruption from
its *location*. Keeping the taxonomy fixed makes incident counts comparable over
time and across chokepoints, and lets users aggregate exposure by mechanism.

## 3. The resilience framework

The resilience profile (`chokepoint_resilience`) decomposes each chokepoint into
distinct, separately-sourced **dimensions**, then combines them into two
transparent composite indices. We follow the standard resilience framing in
which vulnerability rises with what is *at stake* and falls with the *capacity
to absorb* a shock through substitution [@verschuur2025].

### 3.1 Dimensions and definitions

| Dimension | Column(s) | Definition | Primary source |
|---|---|---|---|
| Trade importance | `trade_value_bn_usd`, `oil_transit_mbd`, `oil_share_world_pct` | Annual maritime trade value routed through the chokepoint; crude + petroleum-liquids transit and its share of world seaborne oil | Verschuur & Hall (2025); U.S. EIA |
| Dependency | `n_dependent_countries`, `max_dependency` | Number of countries routing >25% of maritime trade value through it; the single highest country dependency share | Verschuur & Hall (2025) |
| Systemic risk | `evtd_bn_usd` | Expected value of trade disrupted -- the modelled economic loss from disruption | Verschuur & Hall (2025) |
| Network centrality | `degree_centrality`, `strength_weighted`, `betweenness_centrality`, `betweenness_weighted`, `harmonic_centrality`, `pagerank` | Six graph-theory indicators of the chokepoint's position in the country-chokepoint dependency network (Section 4) | computed by the package |
| Redundancy | `redundancy_score`, `has_alternative`, `alt_route`, `extra_days`, `bypass_capacity_mbd` | Whether a viable alternative route exists and its cost in extra time / bypass capacity | compiled from EIA, IEA, UNCTAD |
| Realised disruption | `n_incidents` | Documented incidents in `chokepoint_risks` (2018 on) | per-row cited |

### 3.2 Normalisation and component scores

Each raw dimension is rescaled to `[0, 1]` **relative to these eight
chokepoints** with min-max normalisation,
$\tilde{x}_i = (x_i - \min_j x_j)/(\max_j x_j - \min_j x_j)$, so scores are
*comparative*, not absolute. The component scores are:

- `importance_score` = mean of normalised trade value and normalised oil transit
  (oil missing counts as 0, so oil arteries such as Hormuz are not understated);
- `dependency_score` = normalised `n_dependent_countries`;
- `systemic_risk_score` = normalised `evtd_bn_usd`;
- `redundancy_score` = the documented substitutability coding (0-1; higher = more
  alternatives);
- `exposure_score` = mean of `importance_score`, `dependency_score` and
  `systemic_risk_score`.

### 3.3 The composite indices

$$
\text{resilience\_index} = 100 \times \text{redundancy\_score}
$$
$$
\text{vulnerability\_index} = 100 \times \text{exposure\_score} \times (1 - \text{redundancy\_score})
$$

The **resilience index** is high when a chokepoint can be bypassed; the
**vulnerability index** is high when the stakes are large *and* alternatives are
few -- the multiplicative form encodes that abundant stakes are only dangerous
without a fallback.

These indices are a **transparent, equally-weighted default, not a canonical
measure**. Equal weighting is a neutral prior in the absence of a
user's loss function; because every component ships in the data, users can
re-weight or substitute their own aggregation:

```{r reweight, eval = FALSE}
d <- cp_resilience()
# Example: weight systemic risk twice as heavily as the other exposure terms
w_exposure <- with(d, (importance_score + dependency_score + 2 * systemic_risk_score) / 4)
custom_vulnerability <- 100 * w_exposure * (1 - d$redundancy_score)
```

## 4. Network centrality

Trade value and dependency counts describe how *big* a chokepoint is; they do
not capture its role as a *broker* connecting countries. We therefore compute a
panel of six graph-theory indicators on an explicit network, so this dimension
is **reproducible rather than asserted**.

### 4.1 The network

Using the Verschuur & Hall country-level import dependencies, we build an
**undirected, weighted, bipartite network** with one node per country (ISO3) and
one per chokepoint. An edge links a country to a chokepoint when the country
routes at least **10%** of its maritime import value through it, and the **edge
weight is that dependency share** $s_{ci}\in[0.10,1]$. The graph spans **all**
chokepoints in the source (27, not only our 8), so a node's centrality reflects
its role in the whole maritime network before we extract the eight. At the 10%
threshold the network is a **single connected component** (no isolated nodes),
which is what makes the closeness-type measure below well defined.

### 4.2 The six measures

For a chokepoint node $v$ we report two **unweighted** structural measures and
four that **use the dependency-share weights**:

| Column | Weighted? | Definition | Interpretation |
|---|---|---|---|
| `degree_centrality` | no | degree of $v$, rescaled to the busiest chokepoint | breadth of dependence (how many countries) |
| `strength_weighted` | yes | $\sum_i s_{vi}$ over incident edges | dependency *mass* (how much total reliance) |
| `betweenness_centrality` | no | Freeman betweenness [@freeman1977], normalised | brokerage on the plain topology |
| `betweenness_weighted` | yes | betweenness with edge distance $1/s_{vi}$ | brokerage weighted by dependence intensity |
| `harmonic_centrality` | (topology) | $\tfrac{1}{n-1}\sum_{u\ne v} 1/d(v,u)$, normalised [@boldi2014] | closeness/reachability (disconnection-safe) |
| `pagerank` | yes | stationary distribution of a share-weighted random walk [@brin1998] | recursive influence (important countries depend on it) |

`degree_centrality` is rescaled to the busiest chokepoint; `betweenness_*` and
`harmonic_centrality` are igraph-normalised to `[0, 1]`; `pagerank` sums to 1
over **all** nodes, so chokepoint values are small positive numbers;
`strength_weighted` is a raw sum of shares (an "effective number of
fully-dependent countries"). All six ship as columns of `chokepoint_resilience`:

```{r centrality-table}
cols <- c("location", "degree_centrality", "strength_weighted",
          "betweenness_centrality", "betweenness_weighted",
          "harmonic_centrality", "pagerank")
cp <- chokepoint_resilience[order(-chokepoint_resilience$betweenness_weighted), cols]
print(cp, row.names = FALSE)
```

### 4.3 What they reveal

Centrality is analytically distinct from raw importance. The Strait of Hormuz
carries a fifth of the world's oil yet has **low** betweenness and PageRank: it
is an oil *source* feeding a set of destinations, not a broker between many
countries. Gibraltar and Malacca, by contrast, are the top **brokers** -- the
sole gates between ocean basins -- so their removal would fragment the network
most. The weighted measures add what the unweighted ones drop: `degree_centrality`
**saturates** near 1 for the four busiest arteries (almost every trading country
routes >=10% of imports through them -- a real property of a hub-dominated
network), whereas `strength_weighted` and `betweenness_weighted` still separate
them by *how heavily* countries depend on each. That is why we report both the
structural and the weighted view.

### 4.4 The threshold as a parameter, and what we exclude

The 10% edge threshold is a **chosen parameter**. We verified sensitivity by
recomputing every indicator at 0.05, 0.10 and 0.25 and correlating the eight
chokepoints' scores (Spearman rank correlation vs the 0.10 reference):

| Indicator | vs 0.05 | vs 0.25 |
|---|---:|---:|
| `betweenness_weighted` | 1.00 | 0.98 |
| `pagerank` | 1.00 | 0.95 |
| `strength_weighted` | 0.98 | 0.90 |
| `degree_centrality` | 0.76 | 0.88 |
| `harmonic_centrality` | 0.76 | 0.86 |
| `betweenness_centrality` | 0.74 | 0.76 |

The **weighted** rankings are very robust to the threshold; the unweighted
betweenness/degree and harmonic rankings are moderately stable (magnitudes move
more than ranks). Treat the unweighted columns as an **ordinal** signal, and
re-run `data-raw/prep_zenodo.R` with a different threshold to test robustness for
a specific application.

We deliberately **do not ship** eigenvector centrality or raw closeness (both
degenerate or are hard to interpret on a bipartite graph), the clustering
coefficient (identically 0 -- a bipartite graph has no triangles), or k-core
coreness (low discrimination -- most nodes share one $k$ -- and a
threshold-unstable ranking). A computable number is not automatically a
meaningful one; we report only indicators with a defensible interpretation on
*this* network.

## 5. Quantitative context and provenance

`chokepoint_context` gives a readable per-chokepoint profile -- vessel transits,
vessel and cargo mix, trade shares, top users, local economic dependence and
rerouting cost. Two rules protect its use in research:

1. **Normal-year, like-for-like counts.** Transit figures are representative
   *normal-year* values, never a crisis-depressed snapshot; the `transit_basis`
   column states exactly what is counted (all vessels vs deep-draft; single
   strait vs combined straits; reference year). Disruption *magnitudes* are kept
   in `chokepoint_risks`, not blended into the counts. **Always read
   `transit_basis` before comparing counts across chokepoints** -- for example,
   the Malacca figure is the Malacca + Singapore Straits combined and is not
   comparable to a single-strait count.
2. **Per-figure provenance with a confidence flag.** Every figure is traceable
   through `chokepoint_sources` (value, unit, year, basis, source, URL). The
   `confidence` field is `High` (named authority / official statistics),
   `Medium` (reputable reporting or AIS-derived, e.g. IMF PortWatch via trade
   press) or `Low` (secondary aggregator not resolving to a named authority).
   Filter to keep only what your work can stand behind:

```{r sources, eval = FALSE}
cp_sources(min_confidence = "High")
```

Following the package's "facts, not expression" principle, only factual figures
are reproduced, each attributed; no source's copyrighted table or prose is
copied (`DATA-PROVENANCE.md`).

## 6. Limitations

- **Small n.** Eight chokepoints; normalised scores are comparative within this
  set and should not be read as absolute risk levels.
- **Composite indices are conventions.** Equal weighting is a default, not a
  claim; re-weight for your loss function.
- **Vintage.** Trade and dependency measures reflect the Verschuur & Hall
  base year; oil transit is the latest EIA release; context figures carry their
  own `year`. These do not all share one reference year.
- **Context breadth vs depth.** Some context cells are `NA` or `Low` confidence
  where no authoritative normal-year figure was found; this is deliberate -- the
  package does not pad gaps with unsourced numbers.
- **Centrality is model-dependent.** It depends on the graph definition and the
  10% threshold (Section 4.4); it is one lens, not the truth.

## 7. Reproducibility

The bundled `.rda` objects are rebuilt from scripts in `data-raw/`:
`prep_zenodo.R` (downloads the Verschuur & Hall data, aggregates dependencies and
computes centrality), `prep_context.R` (the curated context and source tables)
and `build_data.R` (assembles and writes all datasets). Re-running them
regenerates the data exactly.

## References
