---
title: "Getting started with chokepointR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started with chokepointR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2,
  dpi = 120, out.width = "100%"
)
library(chokepointR)
has_viz     <- requireNamespace("ggplot2", quietly = TRUE)
has_leaflet <- requireNamespace("leaflet", quietly = TRUE)
cat_cols <- c("Weather and climate risk" = "#0072B2",
              "Security and conflict risk" = "#D55E00",
              "Political and institutional risk" = "#009E73")
```

## What is `chokepointR`?

A large share of seaborne trade — and most seaborne oil — must pass through a
handful of maritime **chokepoints**: canals such as Suez and Panama, and straits
such as Hormuz, Malacca, Bab el-Mandeb, Gibraltar, the Turkish Straits and
Dover. When one is disrupted, the shock propagates through global value chains.

`chokepointR` profiles the **resilience** of these 8 chokepoints. It answers:

> *How much does world trade depend on each chokepoint, how substitutable is it,
> what has actually gone wrong there — and how vulnerable does that make it?*

It is not a measure of global value chains themselves; it supplies the
**risk / resilience layer** that supply-chain and GVC analysis needs.

## Dataset at a glance

```{r}
library(chokepointR)
c(resilience = nrow(chokepoint_resilience),
  incidents  = nrow(chokepoint_risks),
  chokepoints = nrow(chokepoints),
  risk_types  = nrow(risk_types))

# incident coverage and period
range(as.integer(substr(chokepoint_risks$incident_year, 1, 4)))
table(chokepoint_risks$risk_category)
```

## The resilience profile

`cp_resilience()` returns one row per chokepoint with importance, dependency,
systemic-risk and redundancy measures, plus two composite indices.

```{r}
cp_resilience()[, c("location", "trade_value_bn_usd", "n_dependent_countries",
                    "evtd_bn_usd", "resilience_index", "vulnerability_index")]
```

**How the indices are built (transparent, equally-weighted default):**

* `importance_score` blends normalised maritime trade value and oil transit.
* `exposure_score` = mean of importance, dependency and systemic-risk scores.
* `resilience_index` = `100 * redundancy_score` (route substitutability).
* `vulnerability_index` = `100 * exposure_score * (1 - redundancy_score)` — high
  stakes with few alternatives.

All component scores ship in the data, so you can re-weight them for your own
definition of resilience.

```{r vulnerability, eval = has_viz, echo = FALSE, fig.height = 3.8}
library(ggplot2)
d <- cp_resilience()
ggplot(d, aes(reorder(location, vulnerability_index), vulnerability_index,
              fill = resilience_index)) +
  geom_col(width = 0.72) +
  coord_flip() +
  scale_fill_gradient(low = "#d7301f", high = "#2ca25f", name = "Resilience") +
  labs(title = "Chokepoint vulnerability", x = NULL,
       y = "Vulnerability index (0-100)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank())
```

Plotting exposure against redundancy shows the two forces behind the index —
the top-left (high exposure, low redundancy) is where systemic risk concentrates.

```{r scatter, eval = has_viz, echo = FALSE, fig.height = 4}
ggplot(d, aes(redundancy_score, exposure_score)) +
  geom_point(aes(size = trade_value_bn_usd), colour = "#0072B2", alpha = 0.7) +
  geom_text(aes(label = location), size = 3, vjust = -1.1, colour = "grey25") +
  scale_size_area(max_size = 11, guide = "none") +
  scale_x_continuous(expand = expansion(mult = 0.15)) +
  scale_y_continuous(expand = expansion(mult = 0.12)) +
  labs(title = "Exposure vs. route redundancy",
       x = "Redundancy (route substitutability)", y = "Exposure") +
  theme_minimal(base_size = 11)
```

## Explore the incidents

The `chokepoint_risks` log records documented disruptions. Search in plain
language:

```{r}
cp_search("drought Panama")[, c("location", "incident_year", "incident")]
```

Filter on any dimension with `cp_data()`:

```{r}
cp_data(locations = "Strait of Bab el-Mandeb")[, c("incident_year", "risk", "incident")]
```

Map them in one call (interactive **leaflet**, coloured by risk level):

```{r map, eval = has_leaflet, screenshot.force = FALSE}
cp_map(category = "Security and conflict risk")
```

## Reference: the risk taxonomy

```{r, echo = FALSE, eval = requireNamespace("kableExtra", quietly=TRUE)}
library(kableExtra)
kbl(cp_risk(), col.names = c("Risk", "Risk Category", "Risk Code")) |>
  kable_material()
```

## Live hazard signals (optional)

`cp_signals()` fetches recent natural-hazard events (public-domain USGS
earthquake data) near a chokepoint. It needs an internet connection and returns
`NULL` when offline.

```{r, eval = FALSE}
cp_signals("Strait of Hormuz", days = 90, min_magnitude = 4.5)
```

## Data, provenance and citation

`chokepointR` ships **facts** with attribution. Trade-dependency and
systemic-risk measures build on Verschuur & Hall (2025, *Nature Communications*;
data CC BY 4.0); oil transit is from the U.S. EIA (public domain); incidents
carry per-row citations. See `DATA-PROVENANCE.md` for full notes.

```{r}
citation("chokepointR")
```
